---
title: "Sea_Ice_Extent"
author: "Hugh Cipparone, Maggie O'Shea, Nannaphat Sirison"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

## INTRODUCTION

Introduce project objectives here. 

```{r setup, include=FALSE}
#Load in required load packages
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(kableExtra)
library(zoo)
```

## EXPLORING THE DATA

```{r, include= FALSE}
#Loading in the data set
#daily data for Northern hemisphere
df_N<-read.csv("/Users/maggieoshea/Desktop/Spring 2023/TimeSeries/TimeSeriesAnalysis_Sp23/final_project/Data/Daily/N_seaice_extent_daily_v3.0.csv")

#daily date for Southern hemisphere
df_S <- read.csv("/Users/maggieoshea/Desktop/Spring 2023/TimeSeries/TimeSeriesAnalysis_Sp23/final_project/Data/Daily/S_seaice_extent_daily_v3.0.csv")
```

```{r, include = FALSE}
#DATA WRANGLING 
#First row of dataframe are units (extent = 10^6 sq km)
df_N <- df_N[-1,]
df_S <- df_S[-1,]

#Remove spaces in Month and Day columns 
df_N$Month <- gsub("  ", "", df_N$Month)
df_S$Month <- gsub("  ", "", df_S$Month)

df_N$Day <- gsub(" ", "", df_N$Day)
df_S$Day <- gsub(" ", "", df_S$Day)


#Combining date columns into one single column 
df_N$date <- paste(df_N$Year, df_N$Month, df_N$Day, sep = "-")
df_S$date <- paste(df_S$Year, df_S$Month, df_S$Day, sep = "-")

#Convert combined_date column to datetime object 
df_N$date <- ymd(df_N$date)
df_S$date <- ymd(df_S$date)

#Making extent a numeric type
df_N$Extent <- as.numeric(df_N$Extent)
df_S$Extent <- as.numeric(df_S$Extent)

#Remove columns that are not of interest: Keeping only date and extent column
df_N <- df_N[, c("date","Extent")]
df_S <- df_S[, c("date","Extent")]

#subset the data to have only data from 1990 onwards
df_N_subset<-df_N %>% 
  filter(year(date)>=1990)

df_S_subset<-df_S %>% 
  filter(year(date)>=1990)

```

```{r, include = FALSE}
#Converting to time series object 
ts_N <- msts(df_N_subset[,2], start = c(1990,01,01), seasonal.periods = c(7,365.25))
ts_S <- msts(df_S_subset[,2],start = c(1990,01,01), seasonal.periods = c(7,365.23))


```


```{r, echo = FALSE}
ggplot()+
  geom_line(data = df_N, aes(x= date,  y= Extent, color = "North"))+
  geom_line(data= df_S, aes(x=date,  y=Extent, color = "South"))+
  scale_color_manual(values = c("dark blue", "orange")) +
  labs(title = "Sea Ice Extent", y = "Extent: 10^6 sq km", x = "Year")+
  scale_x_date(date_breaks = "10 years", date_labels = "%Y")+
  theme_minimal()
```

Quick discussion of what we see.
This plot of sea ice extent shows a few notable patterns. First, in both north and south sea ice, there is clearly a seasonal patterning which is expected as melting occurs in the warm season while ice growth occurs in the cold season. Importantly, these seasons are the opposite in the north and south. In addition, it shows that the South has greater variation - with more ice in the cold season and less in the warm season. These are all expected patterns, but their representation in the data is important to note. Finally, it appears that there may be a declining trend, particularly visible in the northern sea ice data, however, further analysis is needed to fully examine and understand this. 

## DATA ANALYSIS

#Examining Seasonality and Trend (Decomposition)

```{r, echo = FALSE}
ts_N %>% mstl() %>%autoplot()
```
Discussion of what we see. → Compare seasonality observed weekly vs. monthly

```{r, echo = FALSE}
ts_S %>% mstl() %>%autoplot()


```
Discussion of what we see. → Compare seasonality observed weekly vs. monthly


## FORECASTING

#Model 1 : SARIMA

As we saw in the decomposition, there is no strong seasonality that we observe weekly, but strong monthly seasonality observed. As a result, even though the original day is provided daily, we created a new dataframe that takes the average across days of the month to arrive at a monthly average. This allows us to use the SARIMA forecast. 

```{r, include = FALSE}
#Take mean of sea ice extent across days of the month --> arrive at monthly df 
#1.convert date column to yearmon format
df_N_monthly <- df_N_subset
df_N_monthly$date <- as.yearmon(df_N_monthly$date)

df_S_monthly <- df_S_subset
df_S_monthly$date <- as.yearmon(df_S_monthly$date)

#2.calculate monthly mean
df_N_monthly_mean <- aggregate(Extent ~ date, data = df_N_monthly, FUN = mean)
df_S_monthly_mean <- aggregate(Extent ~ date, data = df_S_monthly, FUN = mean)

#3.Convert to ts object for monthly averaged data
n_obs_n = nrow(df_N_monthly_mean)
#This value (n_for) is 12 months which represents the amount of time subtracted from the training dataset that will then be forecasted and compared
n_for = 12
n_obs_s = nrow(df_S_monthly_mean)

#Training Datasets
ts_N_monthly <- ts(df_N_monthly_mean[1:(n_obs_n-n_for),2], frequency = 12, start= c(1990,1))
ts_S_monthly <- ts(df_S_monthly_mean[1:(n_obs_s-n_for),2], frequency = 12,start= c(1990,1))

#Full datasets
ts_N_monthly_full <- ts(df_N_monthly_mean[,2], frequency = 12, start= c(1990,1))
ts_S_monthly_full <- ts(df_S_monthly_mean[,2], frequency = 12, start= c(1990,1))
```

```{r,echo=TRUE, message=FALSE, warning=FALSE}
#SARIMA Model (autofit) and forecast with training data

#Northern hemisphere model
N_SARIMA_autofit <- auto.arima(ts_N_monthly)
N_SARIMA_for <- forecast(N_SARIMA_autofit, h=12)

#Southern hemisphere model
S_SARIMA_autofit <- auto.arima(ts_S_monthly)
S_SARIMA_for <- forecast(S_SARIMA_autofit, h=12)
```

```{r, echo= FALSE}
#Comparison plot with original full data: Northern hemisphere
autoplot(ts_N_monthly_full) +
autolayer(N_SARIMA_for,series="Seasonal ARIMA",PI=FALSE) +
ylab("Sea ice extent (10^6 sq km)") +
xlab("Time/ Year") +
labs(title="Northern Hemisphere: SARIMA Forecast vs. Monthly Averaged Original Data")+
theme_minimal()
```
Discussion 

```{r, echo= FALSE}
#Comparison plot with original data: Northern hemisphere
autoplot(ts_S_monthly_full, size=0.75) +
autolayer(S_SARIMA_for,series="Seasonal ARIMA",PI=FALSE, size=0.5) +
ylab("Sea ice extent (10^6 sq km)") +
xlab("Time/ Year") +
labs(title="Southern Hemisphere: SARIMA Forecast vs. Monthly Averaged Original Data")+
theme_minimal()
```
Discussion 

## SARIMA Forecast with full datasets
```{r,echo=TRUE, message=FALSE, warning=FALSE}
#SARIMA Model (autofit) and forecast with training data

#Northern hemisphere model
N_SARIMA_fulldata <- auto.arima(ts_N_monthly_full)
N_SARIMA_for_fulldata <- forecast(N_SARIMA_fulldata, h=(12*7))

#Southern hemisphere model
S_SARIMA_fulldata <- auto.arima(ts_S_monthly_full)
S_SARIMA_for_fulldata <- forecast(S_SARIMA_fulldata, h=(12*7))
```

```{r, echo= FALSE}
#Plot Southern Hemisphere forecast
autoplot(ts_N_monthly_full, size=0.75) +
autolayer(N_SARIMA_for_fulldata, series="Seasonal ARIMA",PI=FALSE, size=0.5) +
ylab("Sea ice extent (10^6 sq km)") +
xlab("Time/ Year") +
labs(title="Northern Hemisphere: SARIMA Forecast of Sea Ice Extent")+
theme_minimal()
```

```{r, echo= FALSE}
#Plot Southern Hemisphere forecast
autoplot(ts_S_monthly_full, size=0.75) +
autolayer(S_SARIMA_for_fulldata, series="Seasonal ARIMA",PI=FALSE, size=0.5) +
ylab("Sea ice extent (10^6 sq km)") +
xlab("Time/ Year") +
labs(title="Southern Hemisphere: SARIMA Forecast of Sea Ice Extent")+
theme_minimal()
```



#Model 2 : ETS+STL

## Training Data
```{r}
#Northern Hemisphere
#Fit and forecast STL + ETS model to training data
ETS_trainingdata_N <-  stlf(ts_N_monthly,h=12)

#Plot foresting results
autoplot(ETS_trainingdata_N) + ylab("Active Power")

#Plot model + observed data
autoplot(ts_N_monthly_full) +
  autolayer(ETS_trainingdata_N, series="STL + ETS",PI=FALSE) +
  ylab("Sea ice extent (10^6 sq km)") +
  theme_minimal()

accuracy(ETS_trainingdata_N)
```
Discussion

```{r}
#Southern Hemisphere
#Fit and forecast STL + ETS model to data
ETS_trainingdata_S <-  stlf(ts_S_monthly,h=12)

#Plot foresting results
autoplot(ETS_trainingdata_S) + ylab("Active Power")

#Plot model + observed data
autoplot(ts_S_monthly_full) +
  autolayer(ETS_trainingdata_S, series="STL + ETS",PI=FALSE) +
  ylab("Sea ice extent (10^6 sq km)") +
  theme_minimal()

accuracy(ETS_trainingdata_S)
```

Discussion

Discussion

#Model 3 : State Space Model (BSM)


##Training Data - North
```{r}
BSM_training_N <- StructTS(ts_N_monthly,
                 type="BSM",fixed=c(NA,NA,NA,NA)) 

BSM_training_N_for <- forecast(BSM_training_N,h=n_for)
plot(BSM_training_N_for)

#Plot model + observed data
autoplot(ts_N_monthly_full) +
  autolayer(BSM_training_N_for, series="BSM",PI=FALSE) +
  ylab("Active Power")

accuracy(BSM_training_N_for)

```


##Training Data - South
```{r}
BSM_training_S <- StructTS(ts_S_monthly,
                    type="BSM",fixed=c(NA,NA,NA,NA)) 

BSM_training_S_for <- forecast(BSM_training_S,h=n_for)
plot(BSM_training_S_for)

#Plot model + observed data
autoplot(ts_S_monthly_full) +
  autolayer(BSM_training_S_for, series="BSM",PI=FALSE) +
  ylab("Active Power")

accuracy(BSM_training_S_for)

```



##FORECAST PERFORMANCE AND FORECAST MODEL COMPARISON

#Model 1: SARIMA Accuracy
```{r, include = FALSE}
#Checking SARIMA forecast accuracy 
observed_N <- df_N_monthly_mean[(n_obs_n-n_for+1):n_obs_n,2]
observed_S <- df_S_monthly_mean[(n_obs_s-n_for+1):n_obs_s,2]

N_SARIMA_for_score <- accuracy(N_SARIMA_for$mean,observed_N)
S_SARIMA_for_score<- accuracy(S_SARIMA_for$mean,observed_S)
```


#Model 2: XXX Accuracy

```{r, include = FALSE}
#Checking  forecast accuracy 
ETSforecast_N_score <- accuracy(ETS_trainingdata_N$mean, observed_N)
ETSforecast_S_score<- accuracy(ETS_trainingdata_S$mean, observed_S)
```


#Model 3: BSM Accuracy

```{r}
#Checking BSM Accuracy

N_BSM_for_score <- accuracy(BSM_training_N_for$mean,observed_N)
S_BSM_for_score<- accuracy(BSM_training_S_for$mean,observed_S)
```


#Model 3: XXX Accuracy



## Full Accuracy Table

## Full Accuracy Table
```{r}
#create data frame
N_scores <- as.data.frame(rbind(N_SARIMA_for_score, ETSforecast_N_score, N_BSM_for_score))
S_scores <- as.data.frame(rbind(S_SARIMA_for_score, ETSforecast_S_score, S_BSM_for_score))

row.names(N_scores) <- c("SARIMA", "ETS-STL", "BSM")
row.names(S_scores) <- c("SARIMA", "ETS-STL", "BSM")


```


```{r}

kbl(S_scores, 
      caption = "Forecast Accuracy for Seasonal Monthly Data: South",
      digits = array(5,ncol(S_scores))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(S_scores[,"RMSE"]))

kbl(N_scores, 
      caption = "Forecast Accuracy for Seasonal Monthly Data: North",
      digits = array(5,ncol(N_scores))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(N_scores[,"RMSE"]))
```

##  SELECTED FULL MODEL 
## Full Dataset(s)
```{r}
#Northern Hemisphere
#Fit and forecast STL + ETS model to training data
ETS_full_N <-  stlf(ts_N_monthly_full,h=(12*7))

#Plot foresting results
autoplot(ETS_full_N) + ylab("Active Power")

#Plot model + observed data
autoplot(ts_N_monthly_full) +
  autolayer(ETS_full_N, series="STL + ETS",PI=FALSE) +
  ylab("Sea ice extent (10^6 sq km)") +
  theme_minimal()

```



```{r}
#Southern Hemisphere
#Fit and forecast STL + ETS model to data
ETS_full_S <-  stlf(ts_S_monthly_full,h=(12*7))

#Plot foresting results
autoplot(ETS_full_S) + ylab("Active Power")

#Plot model + observed data
autoplot(ts_S_monthly_full) +
  autolayer(ETS_full_S, series="STL + ETS",PI=FALSE) +
  ylab("Sea ice extent (10^6 sq km)") +
  theme_minimal()

```


Discussion

##DATA SOURCE





## CODE DUMP


