---
title: "Sea_Ice_Extent"
author: "Hugh Cipparone, Maggie O'Shea, Nannaphat Sirison"
date: "`r Sys.Date()`"
output: pdf_document
---

## Yearly Models

```{r setup, include=FALSE}
#Load in required load packages
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(kableExtra)
library(zoo)

```

## EXPLORING THE DATA

```{r, include= FALSE}
#Loading in the data set
#daily data for Northern hemisphere
df_N<-read.csv("/Users/maggieoshea/Desktop/Spring 2023/TimeSeries/TimeSeriesAnalysis_Sp23/final_project/Data/Daily/N_seaice_extent_daily_v3.0.csv")

#daily date for Southern hemisphere
df_S <- read.csv("/Users/maggieoshea/Desktop/Spring 2023/TimeSeries/TimeSeriesAnalysis_Sp23/final_project/Data/Daily/S_seaice_extent_daily_v3.0.csv")
```

```{r, include = FALSE}
#DATA WRANGLING 
#First row of dataframe are units (extent = 10^6 sq km)
df_N <- df_N[-1,]
df_S <- df_S[-1,]

#Remove spaces in Month and Day columns 
df_N$Month <- gsub("  ", "", df_N$Month)
df_S$Month <- gsub("  ", "", df_S$Month)

df_N$Day <- gsub(" ", "", df_N$Day)
df_S$Day <- gsub(" ", "", df_S$Day)


#Combining date columns into one single column 
df_N$date <- paste(df_N$Year, df_N$Month, df_N$Day, sep = "-")
df_S$date <- paste(df_S$Year, df_S$Month, df_S$Day, sep = "-")

#Convert combined_date column to datetime object 
df_N$date <- ymd(df_N$date)
df_S$date <- ymd(df_S$date)

#Making extent a numeric type
df_N$Extent <- as.numeric(df_N$Extent)
df_S$Extent <- as.numeric(df_S$Extent)

#Remove columns that are not of interest: Keeping only date and extent column
df_N <- df_N[, c("date","Extent")]
df_S <- df_S[, c("date","Extent")]
```


```{r, include = FALSE}
#Take mean of sea ice extent across days of the month --> arrive at monthly df 
#1.convert date column to yearmon format
df_N_yearly <- df_N
df_N_yearly$date <- as.numeric(format(df_N_yearly$date,'%Y'))

df_S_yearly <- df_S
df_S_yearly$date <- as.numeric(format(df_S_yearly$date,'%Y'))

#2.calculate monthly mean
#Need to remove 2023 because the year isn't over so the south will appear to have plummeted on mean sea ice while the north will seem like it's grown because they haven't had both their seasons yet. 
df_N_annual_mean <- aggregate(Extent ~ date, data = df_N_yearly, FUN = mean)%>%
  filter(date!=2023)
df_S_annual_mean <- aggregate(Extent ~ date, data = df_S_yearly, FUN = mean)%>%
  filter(date!=2023)

#3.Convert to ts object for monthly averaged data
n_obs_n = nrow(df_N_annual_mean)
#This value (n_for) is 5 years which represents the amount of time subtracted from the training dataset that will then be forecasted and compared
n_for = 5
n_obs_s = nrow(df_S_annual_mean)

#Training Datasets
ts_N_annual <- ts(df_N_annual_mean[1:(n_obs_n-n_for),2], frequency = 1, start= c(1978))
ts_S_annual <- ts(df_S_annual_mean[1:(n_obs_s-n_for),2], frequency = 1,start= c(1978))

#Full datasets
ts_N_annual_full <- ts(df_N_annual_mean[,2], frequency = 1, start= c(1978))
ts_S_annual_full <- ts(df_S_annual_mean[,2], frequency = 1, start= c(1978))
```


```{r, echo = FALSE}
ggplot()+
  geom_line(data = df_N_annual_mean, aes(x= date,  y= Extent, color = "North"))+
  geom_line(data= df_S_annual_mean, aes(x=date,  y=Extent, color = "South"))+
  scale_color_manual(values = c("dark blue", "orange")) +
  labs(title = "Sea Ice Extent", y = "Extent: 10^6 sq km", x = "Year")+
  #scale_x_date(date_breaks = "10 years", date_labels = "%Y")+
  theme_minimal()
```

```{r}
par(mfrow=c(1,2))
  acf(ts_N_annual_full)
  pacf(ts_N_annual_full)

par(mfrow=c(1,2))
  acf(ts_S_annual_full)
  pacf(ts_S_annual_full)
```



# ARIMA on Annual Data

```{r,echo=TRUE, message=FALSE, warning=FALSE}
#ARIMA Model (autofit) and forecast with training data

#Northern hemisphere model
N_ARIMA_training <- auto.arima(ts_N_annual)
N_ARIMA_for_train <- forecast(N_ARIMA_training, h=5)

#Southern hemisphere model
S_ARIMA_training <- auto.arima(ts_S_annual)
S_ARIMA_for_train <- forecast(S_ARIMA_training, h=5)
```


```{r, echo= FALSE}
#Comparison plot with original full data: Northern hemisphere
autoplot(ts_N_annual_full) +
autolayer(N_ARIMA_for_train,series="ARIMA",PI=FALSE) +
ylab("Sea ice extent (10^6 sq km)") +
xlab("Time/ Year") +
labs(title="Northern Hemisphere: SARIMA Forecast vs. Monthly Averaged Original Data")+
theme_minimal()

autoplot(ts_S_annual_full) +
autolayer(S_ARIMA_for_train,series="ARIMA",PI=FALSE) +
ylab("Sea ice extent (10^6 sq km)") +
xlab("Time/ Year") +
labs(title="Northern Hemisphere: SARIMA Forecast vs. Monthly Averaged Original Data")+
theme_minimal()
```

## ARIMA on full dataset
```{r,echo=TRUE, message=FALSE, warning=FALSE}
#ARIMA Model (autofit) and forecast with *full* dataset

#Northern hemisphere model
N_ARIMA_full <- auto.arima(ts_N_annual_full)
N_ARIMA_for_full <- forecast(N_ARIMA_full, h=10)

#Southern hemisphere model
S_ARIMA_full <- auto.arima(ts_S_annual_full)
S_ARIMA_for_full <- forecast(S_ARIMA_full, h=10)
```


```{r, echo= FALSE}
#Comparison plot with original full data: Northern hemisphere
autoplot(ts_N_annual_full) +
autolayer(N_ARIMA_for_full,series="ARIMA",PI=FALSE) +
ylab("Sea ice extent (10^6 sq km)") +
xlab("Time/ Year") +
labs(title="Northern Hemisphere: SARIMA Forecast vs. Monthly Averaged Original Data")+
theme_minimal()

autoplot(ts_S_annual_full) +
autolayer(S_ARIMA_for_full,series="ARIMA",PI=FALSE) +
ylab("Sea ice extent (10^6 sq km)") +
xlab("Time/ Year") +
labs(title="Northern Hemisphere: SARIMA Forecast vs. Monthly Averaged Original Data")+
theme_minimal()
```

You can see in the Northern Hemisphere plots that the model predicts a steady decline in sea ice, ultimately approaching zero sometime after 2050. However, we know that the decline in sea ice is exponential, not linear due to albedo leading to a positive feedback loop of melting. Because of this, scientists predict that the arctic will have no summer sea ice as soon as 2035 which is much sooner than our model predicts. 



