---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 3 - Due date 02/10/23"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---
```{r}

```
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
library(tidyverse)

#Import Data
library(readxl)
energy_data<- read_excel("../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", skip=12, sheet="Monthly Data", col_names=FALSE)
read_col_names <- read_excel(path="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 
colnames(energy_data) <- read_col_names

energy_clean <- energy_data%>%
  select("Month", "Total Biomass Energy Production", "Total Renewable Energy Production", "Hydroelectric Power Consumption")

ts_energy <- ts(energy_clean[,2:4], frequency = 12, start = c(1973, 1))

```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)

```{r}

for(i in 1:3){
  par(mfrow = c(1,3))
  plot(ts_energy[,i],type="l",col="blue")
  abline(h=mean(ts_energy[,i]),col="red")
  acf(energy_clean[,i+1], lag.max = 40)
  pacf(energy_clean[,i+1], lag.max = 40)
}

```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

#### Q2 Answer: 
The plot for biomass energy production does appear to have a trend. The ACF for both Biomass Energy production and Total Renewable Energy production both show a deterministic trend in the ACF plot which show a high correlation in the early lags that declines consistently. The hydroelectric power plot shows a clear seasonal component that possibly declines as well, though further analysis is needed. 

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.
 
#### Biomass
```{r}
#Biomass

nobs <- nrow(energy_clean) 
t <- 1:nobs 

linearTrend_biomass <- lm(energy_clean$`Total Biomass Energy Production` ~ t)
summary(linearTrend_biomass)

beta0_biomass <- linearTrend_biomass$coefficients[1]
beta1_biomass <- linearTrend_biomass$coefficients[2]
```

##### Interpret Biomass Linear Trend:
The trend helps to identify the long-term tendency of the data, in this case of Total Biomass Production. The positive coefficients (and statistical significance), both for the intercept and time, indicate that the for ever increase in time (t), there is an associated increase in biomass production. The intercept is the value of Total Biomass production at time 0, and thus the rest of the values of biomass production (at time 1 and on) are in relation to this value of 133.7. Because this and beta1 are positive, the trend is a positive trend with biomass production increasing over time. 

#### Renewable 
```{r}
#Renewable
linearTrend_renew<- lm(energy_clean$`Total Renewable Energy Production` ~ t)
summary(linearTrend_renew)

beta0_renew <- linearTrend_renew$coefficients[1]
beta1_renew <- linearTrend_renew$coefficients[2]
```

##### Interpret Renewable Energy Linear Trend:
As in the Biomass linear model, time and the intercept are statistically significant and have positive coefficients. The intercept indicates that at time 0 the renewable energy production would be 312.25 and increases from there - an increasing trend over time. 

#### Hydroelectric
```{r}
linearTrend_hydro <- lm(energy_clean$`Hydroelectric Power Consumption` ~ t)
summary(linearTrend_hydro)

beta0_hydro <- linearTrend_hydro$coefficients[1]
beta1_hydro <- linearTrend_hydro$coefficients[2]
```

##### Interpret Hydroelectric Power Linear Trend:
Again the intercept and t are statistically significant for hydroelectic power consumption. However, beta-1 for hydroelectric power is negative. The intercept is at 259.898 which is the value of hydroelectric power at time 0, but because the coefficient on time is negative (-0.083), this declines over time. 
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r}
#Biomass
detrend_biomass <- energy_clean$`Total Biomass Energy Production` - (beta0_biomass+beta1_biomass*t)

ggplot(energy_clean, aes(x=Month, y=`Total Biomass Energy Production`))+
  geom_line(color = "blue")+
  geom_line(aes(y=detrend_biomass), color = "green")

#Renewable Energy
detrend_RE <- energy_clean$`Total Renewable Energy Production` - (beta0_renew+beta1_renew*t)

ggplot(energy_clean, aes(x=Month, y=`Total Renewable Energy Production`))+
  geom_line(color = "blue")+
  geom_line(aes(y=detrend_RE), color = "green")

#Hydroelectric Power
detrend_hydro <- energy_clean$`Hydroelectric Power Consumption`- (beta0_hydro+beta1_hydro*t)

ggplot(energy_clean, aes(x=Month, y=`Hydroelectric Power Consumption`))+
  geom_line(color = "blue")+
  geom_line(aes(y=detrend_hydro), color = "green")

```

#### Q4 Answer:
After removing the trend and plotting the detrended series, all of the plots changed. All of the plots moved down on the graph, with lower values than the non de-trended series. In addition, the first two plots (total renewable production and total biomass production) also were not only lower on the graph but the difference between the start and end values of energy production were more similar in both plots - essentially it is visible that the increasing trend in the data was removed making it closer to a flat trend over time. The decreasing trend in the hydroelectric plot is subtle, and thus this de-trending is more difficult to see in the detrended plot. 

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?

```{r}
#Biomass 

acf(detrend_biomass, lag.max=40)
#acf(energy_clean$`Total Biomass Energy Production`, lag.max = 40)
pacf(detrend_biomass, lag.max=40)
#pacf(energy_clean$`Total Biomass Energy Production`, lag.max = 40)

#Renewable Energy
acf(detrend_RE, lag.max=40)
#acf(energy_clean$`Total Renewable Energy Production`, lag.max = 40)
pacf(detrend_RE, lag.max=40)
#pacf(energy_clean$`Total Renewable Energy Production`, lag.max = 40)


#Hydroelectric
acf(detrend_hydro, lag.max=40)
#acf(energy_clean$`Hydroelectric Power Consumption`, lag.max = 40)

pacf(detrend_hydro, lag.max=40)

```
##### Question 5 Answer: 
In comparing ACF for biomass and for renewable energy, the detrended plots still seem to have a declining trend but the seasonal elements of the data are much more obvious, a change that is especially stark in the biomass production plot. The biomass production PACF and the renewable energy PACF are similar in that where there are high correlations in the PACF with the trend there are also high correlations in the detrended PACF. However, the detrended PACF correlations seem to be generally higher than the PACF with the trends. Finally, the ACF for hydroelectric without the trend has lower correlations, with more negative correlations as well as just broadly lower correlations even where positive. 

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

#### Q6 Pre-analysis Answer
The hydroelectric consumption series has a clear seasonal trend according to the ACF. The other series, biomass and renewables, there appears to be seasonality as well, but more analysis is needed. 

```{r}
#Biomass
biomass_dummies <- seasonaldummy(ts_energy[,1])

biomass_seasMeans <- lm(energy_clean$`Total Biomass Energy Production` ~biomass_dummies)
summary(biomass_seasMeans )

beta0_biomassS <- biomass_seasMeans $coefficients[1]
beta1_biomassS <- biomass_seasMeans $coefficients[2:12]

biomass_seas_comp <- array(0,nobs)
for (i in 1:nobs)(
  biomass_seas_comp[i] = beta0 + beta1 %*% biomass_dummies[i,]
)

ggplot(energy_clean, aes(x=Month, y=`Total Biomass Energy Production`))+
  geom_line(color="blue")+
  geom_line(aes(y=biomass_seas_comp), color="orange")

#Total Renewable Energy
RE_dummies <- seasonaldummy(ts_energy[,2])

RE_seasMeans <- lm(energy_clean$`Total Renewable Energy Production` ~ RE_dummies)
summary(RE_seasMeans )

beta0_reS <- RE_seasMeans$coefficients[1]
beta1_reS <- RE_seasMeans$coefficients[2:12]

RE_seas_comp <- array(0,nobs)
for (i in 1:nobs)(
  RE_seas_comp[i] = beta0_reS + beta1_reS %*% RE_dummies[i,]
)

ggplot(energy_clean, aes(x=Month, y=`Total Renewable Energy Production`))+
  geom_line(color="blue")+
  geom_line(aes(y=RE_seas_comp), color="orange")

#Hydroelectric
hydro_dummies <- seasonaldummy(ts_energy[,3])

hydro_seasMeans <- lm(energy_clean$`Hydroelectric Power Consumption` ~hydro_dummies)
summary(hydro_seasMeans)

beta0_hydroS <- hydro_seasMeans$coefficients[1]
beta1_hydroS <- hydro_seasMeans$coefficients[2:12]

hydro_seas_comp <- array(0,nobs)
for (i in 1:nobs)(
  hydro_seas_comp[i] = beta0_hydroS + beta1_hydroS %*% hydro_dummies[i,]
)

ggplot(energy_clean, aes(x=Month, y=`Hydroelectric Power Consumption`))+
  geom_line(color="blue")+
  geom_line(aes(y=hydro_seas_comp), color="orange")
```


### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

```{r}
detrend_Seas_biomass <- energy_clean$`Total Biomass Energy Production` - (beta0_biomassS+beta1_biomassS*t)
detrend_Seas_RE <- energy_clean$`Total Renewable Energy Production` - (beta0_reS+beta1_reS*t)
detrend_Seas_hydro <- energy_clean$`Hydroelectric Power Consumption`- (beta0_hydroS+beta1_hydroS*t)

ggplot(energy_clean, aes(x=Month, y=`Total Biomass Energy Production`))+
  geom_line(color="blue")+
  geom_line(aes(y=detrend_Seas_biomass), color="orange")

ggplot(energy_clean, aes(x=Month, y=`Total Renewable Energy Production`))+
  geom_line(color="blue")+
  geom_line(aes(y=detrend_Seas_RE), color="orange")

ggplot(energy_clean, aes(x=Month, y=`Hydroelectric Power Consumption`))+
  geom_line(color="blue")+
  geom_line(aes(y=detrend_Seas_hydro), color="orange")



```


### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?

```{r}
#Biomass 
acf(detrend_Seas_biomass, lag.max=40)
#acf(energy_clean$`Total Biomass Energy Production`, lag.max = 40)
pacf(detrend_Seas_biomass, lag.max=40)
#pacf(energy_clean$`Total Biomass Energy Production`, lag.max = 40)

#Renewable Energy
acf(detrend_Seas_RE, lag.max=40)
#acf(energy_clean$`Total Renewable Energy Production`, lag.max = 40)
pacf(detrend_Seas_RE, lag.max=40)
#pacf(energy_clean$`Total Renewable Energy Production`, lag.max = 40)


#Hydroelectric
acf(detrend_Seas_hydro, lag.max=40)
#acf(energy_clean$`Hydroelectric Power Consumption`, lag.max = 40)
pacf(detrend_Seas_hydro, lag.max=40)
#pacf(energy_clean$`Hydroelectric Power Consumption`, lag.max = 40)

```

